\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-27 rebuttal.tex

\reply{Reinhold Scherer}
Major comments:
\begin{enumerate}
	\item Throughout the text, qualitative statements have been quantified where possible and appropriate.
	\item  All mathematical equations and derivations have thoroughly
	been checked for correctness and notational consistency, and have been
	corrected where necessary.
	\item As indicated by the reviewer, true on-line experiments with live
	feedback have indeed not been reported on in this thesis,
  but were foreseen to take	place within the framework of this doctoral study.
	I agree that these experiments would be valuable to supplement the
	presented conclusions and are the logical next step.
	Indeed, the start of experimentation with healthy control participants,
	as well as with patient groups, was heavily delayed by the COVID-19 pandemic, as the
	preparatory research started in September 2019 and I officially
	enrolled in the doctoral training program in January 2020.
	This restricted not only our access to experiment subjects for a
	significant period at the start of the PhD -- and even longer for
	vulnerable patient groups -- but also shifted priorities for the medical
	staff at our partner institutions and patient centers.
	Hence, work early in the PhD mainly focused on the development of
	computational methods.
	We are currently starting up true on-line experimentation in cooperation
	with our partner patient centers, with an adapted interface and experimental protocol.
  This limitation has now clearly been mentioned in section 8.2.
\end{enumerate}
Minor comments:
\begin{enumerate}
	\item Recorded datasets could not be made publicly available since the
	ethical protocol (Ethics Commission of University Hospital Leuven,
	S62547 v1.4) for the conducted studies did not include provisions
	publishing the data.
\end{enumerate}

\reply{Fabien Lotte}
Major comments:
\begin{enumerate}
	\item
    Part of the confusion between active and passive BCI might be due
	  to the poor choice of phrasing ``paradigms that require little active
    participation from the user, like redirecting attention or initiating
    imagined actions, are classified as passive BCI.''
	  This has been rewritten to ``in contrast to paradigms that require the
    user's active participation, for
	  instance through redirecting attention or initiating imagined actions,
	  paradigms free of this requirement are classified as passive BCI,''
	  for clarity.
	  Neurofeedback has been reclassified as active BCI in the text and in
	  figure 1.3.
	\item I agree that investigating the effect of shrinkage regularization on the
    performance of XDAWN+RG is necessary to obtain a fair comparison method.
    This is especially in the case of interest here, when data are scarcely
    available and the proposed methods also rely on shrinkage.
    While setting up the decoding performance comparison experiment, we
    evaluated performance with and without Ledoit-Wolf shrinkage for both the
    estimation of the XDAWN filters and the subsequent covariance matrices.
    Indeed, we expected that this regularization would increase performance,
    just as in the STBF case. However, we noticed that highest accuracy for the
    XDAWN+RG method was reached without shrinkage in the XDAWN filter
    estimation and the covariance estimation, and when retaining a sample rate
    of 2048 Hz.
    This has now clearly been indicated in the manuscript.
  \item
    As indicated by the reviewer, however, using a 2048 Hz sample rate for one of the
    classifiers, while 32 Hz has been used for the others, gives a skewed image
    of the training time.
    A fair comparison of training times has now been reported in section 1.3.3
    and figure 1.4 with XDAWN+RG also using a 32 Hz sample rate.
  \item Appendix 1.A now fully reports all obtained accuracies to give a better
    insight in effect sizes. Wherever the significant effects have been mentioned
    in the text, the effect size has now been reported along with it.
	\item
    In the results section of chapter 4, statistical comparisons were
	  only performed between methods for which we evaluated the performance
	  ourselves (HODA, PARAFACDA and BTTDA).
	  Since the MOABB benchmarking framework~\cite{Aristimunha2023} allows for
	  fair comparison of performance metrics reported across different works,
    the analyses for other comparison methods were not re-run by us (as
    reported).
    Instead, but performances were taken from the MOABB benchmark database~\cite{Chevallier2024}.
	  Unfortunately, this source provides only the aggregated performance metrics
	  and not the metrics per subject, session and cross-validation fold,
	  hence we could not perform relevant statistical comparison with these
	  comparison models.
  \item
    Tables 6.2 and 6.3 now report the absolute cross-condition ROC-AUC values
    per model and dataset for a more complete and interpretable presentation of
    the results, and the absolute values have been mentioned when referring to
    effects in the results section.
\end{enumerate}
Minor comments:
\begin{enumerate}
	\item According to the annotations, typographical errors have been
	corrected in text and figures, and imprecise or inaccurate statements have been adjusted.
	The recommended references have been discussed and, where necessary,
	statements were updated accordingly.
  Figure 1.5 was corrected and matched with the correct caption.
\end{enumerate}

\reply{Andrea K\"ubler}
Major comments:
\begin{enumerate}
  \item  I agree that the presented work could have been strengthened by the
  formulation of clearly testable hypotheses, including a general hypothesis
  and per-study subhypotheses.
  I would partially attribute the lack thereof to the exploratory and
  engineering-oriented nature of this thesis.
  The original hypotheses were rephrased as directed research objectives in
  section 2.3, but these have now again explicitly been reformulated to the
  following hypothesis: \emph{``a suited decoding strategy can improve the
    (gaze-independent) accuracy of a spatially organized visual oddball
    ERP-based brain''}.
  The findings regarding this hypothesis have now also been discussed
    explicitly in section 8.1.3.
  \item As mentioned above in the reply to prof. Reinhold Scherer's third major
    comment, this thesis does indeed not report on the foreseen on-line
    experimentation due to several factors.
    Current work focuses on on-line evaluation with user interest groups.
  \item
    A clearer distinction between methods, results and discussion has been made
    in several chapters.
    In chapter 5, the synthetic data generation procedure has been
    moved to the method section as section 5.3.4.
    In chapter 6, the analysis investigating observed jitter per condition has
    been moved out of the discussion section and has been given a place in the
    results section as section 6.3.3.
    Sections in chapter 7 have been renamed to follow a proper IMRaD structure.
    Here, the description of visual motor assessment and the
    discussion of outcomes regard visual skills have been separated and placed
    respectively in the methods and results section.
    Since chapter 4 focuses on methodological contributions which are then
    verified in experiments that are rather independent from these methods, I
    believe it is clearer to keep the presentation of the evaluation methods
    and the results together per data modality in a separate `Experiments'
    section, instead of also moving the description of evaluation methods to
    the method section.
  \item I very much appreciate the instruction of the reviewer to further
    investigate user-centered design (UCD)~\cite{Standardization2009,Kuebler2014}.
    In fact, there was a large overlap between the principles and methods laid
    out in the UCD framework and the topics discussed in section 8.2.
    The UCD standard and its applications to BCI research provide a structured
    approach to discussing the limitations and recommendations arising from this
    work.
    Furthermore, they form a guideline for conducting ongoing and future on-line patient
    studies.
    UCD has now extensively been referenced throughout chapter 8.
\end{enumerate}
Minor comments:
\begin{enumerate}
  \item The description of fMRI and fNIRS has been
    corrected to reflect the properties of the BOLD signal and the fNIRS blood
    oxygenation signal.
  \item	Neurofeedback has been reclassified as active BCI in the text and in
	  figure 1.3.
    The introduction to passive BCIs has been rephrased to focus on
    \emph{control through detection} instead of only \emph{monitoring}.
    Monitoring can be a part of passive BCIs, e.g. for clinical use, but I
    agree that the core focus should lie here indeed on how to use the output
    for useful control.
    Figure 1.3 has been updated accordingly.
  \item A brief review of the state-of-the art of tactile BCIs in the
    context of this thesis has been added to section 2.2.1.
  \item I agree that loss of vision is an important consideration which affects
    the merits of specific BCI paradigms. This has now been discussed in the
    literature review presented in section 2.2.1. The research objectives
    put forward in section 2.3 state now state clearly that our study concerns
    those individuals retaining sufficient vision.
\end{enumerate}

\reply{Maarten De Vos}
Major comments:
\begin{enumerate}
	\item Mathematical equations and derivations have thoroughly
	been checked for correctness and notational consistency, and have been
	corrected throughout the manuscript where necessary.
	The paragraph on Block-Toeplitz linear discriminant analysis (tLDA) in
	section 1.6.4 has also been corrected, and it was expanded for clarity.
	Figures 3.1 and 3.2 have been regenerated and relabeled for consistent
	formatting.
	\item The Wilcoxon signed-rank, a non-parametric,
    paired test was used to compare ROC-AUC values in sections 4.3.3 and 6.3.1.
    I agree that the DeLong test is indeed an appropriate test for
    statistically comparing ROC-AUCs between two models, but it is commonly used
    to compare ROCs obtained from a single test dataset (potentially obtained
    through stacking cross-validated predictions).
    Our experiments, on the other hand, involve calculating ROC-AUCs per
    cross-validation fold, yielding datasets of multiple ROC-AUCs to compare.
    In this case, the Wilcoxon signed-rank test is the appropriate
    test~\cite{Rainio2024}.

    While both approaches are valid strategies and commonly used~\cite{Rainio2024}, I believe
    the latter is more interpretable in the context of the generalizability
    of machine learning models, since it more directly reflects the variance
    across cross-validation folds.
    Hence, I recognize the potential application of the DeLong test here,
    but I would argue that using Wilcoxon signed-rank tests in combination with
    the evaluation strategy employed in the experiments is still a valid approach.
   \end{enumerate}
Minor comments:
\begin{enumerate}
	\item Unsupported claims regarding generalizability have been omitted
	from the abstract.
	\item Throughout the manuscript, different EEG band-pass filtering
	settings have been used.
	Chapter 1 indicates that ERPs are usually band-pass filtered between
	0.5 and 16 Hz.
	These were also the preferred cut-off frequencies for other
	(unpublished) analyses used during the course of the thesis work, an
	those used for the analyses in chapter 3.

	However, ERP datasets evaluated in chapter 4 were band-pass filtered
	between 1 and 24 Hz, and in chapter 6 between 0.1 and 20 Hz.
	While there is no general consensus in the field about the optimal
	cut-off frequencies for P300 detection and ERP analysis in general,
	these values all fall well within the accepted
	bounds~\cite{Bougrain2012}.
	Decisions to deviate from this were made consciously.

	In chapter 3, filtering between 1 and 24 Hz was applied for consistency
	with the recommended ERP preprocessing pipeline of the MOABB
	benchmarking framework~\cite{Aristimunha2023}.
	This allows for a fair comparison under equal circumstances of the
	performance of the proposed algorithm with performances of other
	algorithms reported by the MOABB benchmarking
	database~\cite{Chevallier2024}.

	In chapter 6, the low-pass frequency was increased from 16 Hz to 20 Hz,
	since the analyses in this explicitly involve the short-time effects of
	latencies of ERP components, and the analysis of faster, early ERP
	components. A low low-pass frequency would harm the precision of
	latency effects in the filtered signal.
	A lower high-pass filter of 0.1 Hz was used here to reveal the
	influence of the proposed alignment approach on potential
	slower late components.
	\item Typographical errors have been corrected throughout the
	manuscript.
\end{enumerate}

### 2024-11-27 chapters/acknowledgment.tex

\chapter*{Acknowledgments}
We thank Arno Libert,
Tjaša Mlinarič, Yide Li, Qiang Sun and Mani Mirsaeedi for their help in collecting data
for the works presented in \cref{sec:covert-align} and
\cref{sec:patients}.
We would also like to thank the other colleagues at the KU Leuven Computational
Neurosciences group at KU Leuven and the BCI team at the University of Lille,
for their general support.
Furthermore, we thank the teams at TRAINM Neuro Rehab Clinics Antwerp,
the UZ Leuven Neuromuscular Reference Renter and the Fondation Partage et Vie.
Finally, we wish to acknowledge dr.\ Axel Faes for his inspiration in conceptualizing
the work presented in~\cref{sec:bttda}.
We thank the Flemish Supercomputer Center (VSC) and the High-Performance
Computing (HPC) center of the KU Leuven for allowing us to execute our
computational experiments on their systems.

>>>>

\chapter*{Acknowledgments}
We thank Arno Libert,
Tjaša Mlinarič, Yide Li, Qiang Sun and Mani Mirsaeedi for their help in collecting data
for the works presented in \cref{sec:covert-align} and
\cref{sec:patients}.
We would also like to thank the other colleagues at the KU Leuven Computational
Neurosciences group at KU Leuven and the BCI team at the University of Lille,
for their general support.
Furthermore, we thank the teams at TRAINM Neuro Rehab Clinics Antwerp,
the UZ Leuven Neuromuscular Reference Renter and the Fondation Partage et Vie
specialized care home.
We wish to acknowledge dr.\ Axel Faes for his inspiration in conceptualizing
the work presented in~\cref{sec:bttda}.
Finally, we thank the Flemish Supercomputer Center (VSC) and the High-Performance
Computing (HPC) center of the KU Leuven for allowing us to execute our
computational experiments on their systems.

### 2024-11-27 chapters/bttda.tex
Reply to DeLong test

\cref{tab:mi-score} lists the cross-validated classification scores for
the evaluated motor imagery datasets.
Note that, in line with~\textcite{Chevallier2024}, accuracy is reported for the
multi-class classification problem in BNCI2014-001, while ROC-AUC was reported for the
binary classification problem in BNCI2014-004.
\begin{table}[t]
  \centering
	\footnotesize
	\input{chapters/bttda/score_mi.tex}
  \caption[Within-session classification score for 2 motor imagery datasets.]{Classification score for
		cross-validated within-session evaluation for HODA+LDA and our proposed decoders
		PARAFACDA+LDA and BTTDA+LDA evaluated on 2 motor imagery datasets.
		Scores for other decoders were taken from \textcite{Chevallier2024}. Accuracy
    is listed for BNCI2014-001, and area under the receiver-operator
    characteristic curve for BNCI2014-004. BTTDA outperforms HODA and PARAFACDA
    for BNCI2014-001 but does not reach performance comparable to the current
    state-of-the-art. For BNCI2014-004, PARAFACDA and BTTDA perform
    approximately on par with HODA and the current state-of-the art.}
	\label{tab:mi-score}
\end{table}
For the BNCI2014-001 dataset, HODA+LDA and our proposed decoders do not reach
satisfactory performance compared to the comparison methods, yet both
PARAFACDA+LDA and BTTDA+LDA improve upon HODA+LDA, and for BTTDA+LDA this
difference is significant ($p=0.0269$).
For the BNCI2014-004 dataset, the performance gap with comparison methods is
smaller, but no significant differences were found between HODA+LDA,
PARAFACDA+LDA, and BTTDA+LDA.

@@

Nevertheless, when considering the relative improvement over the original
HODA model, our results show that BTTDA+LDA performs consistently on par or
higher than HODA+LDA, with performances of respectively $86.43>83.25$ and
$85.08>82.57$ for \ac{erp} datasets and $55.13>53.51$ and $80.60<80.88$ for
\ac{mi} datasets.
The classification score for BTTDA+LDA was also slightly
higher than that of
restricted PARAFACDA+LDA for the ERP datasets ($86.43>86.19$ and $85.06>84.85$
respectively) and one of the MI datasets ($55.13>54.34$ and $80.60<80.88$
respectively), but these results were not significant, and further studies with
more datasets and :subjects should show whether this holds.
\Cref{fig:blocks} shows that there is added value in finding extra blocks
over the first HODA block.
While no proof is given, we notice that NMSE monotonically decreases, indicating that
since the training score keeps increasing, this suggests that
eventually all the variation in the signal will be explained by the model
while still extracting features that are maximally discriminant.
Eventually, the number of blocks will reach a point of diminishing validation
score returns, when adding extra features to the decision classifier increases
its risk of overfitting instead of adding extra useful discriminatory
information.
Together with the improved classification scores presented, these results
point to the potential of our more flexible model block-term or
its special PARAFAC-structured case over a Tucker-structured model given proper block and
rank selection.
Since the optimal ranks for HODA+LDA were also determined through
cross-validation, BTTDA+LDA can improve over the first HODA block.
If validation shows that this is not possible, such as is the case for
BNCI2014-004, the BTTDA model is truncated to the first HODA block and little
performance is lost.

>>>>

\cref{tab:mi-score} lists the cross-validated classification scores for
the evaluated motor imagery datasets.
Note that, in line with~\textcite{Chevallier2024}, accuracy is reported for the
multi-class classification problem in BNCI2014-001, while ROC-AUC was reported for the
binary classification problem in BNCI2014-004.
\begin{table}[t]
  \centering
	\footnotesize
	\input{chapters/bttda/score_mi.tex}
  \caption[Within-session classification score for 2 motor imagery datasets.]{Classification score for
		cross-validated within-session evaluation for HODA+LDA and our proposed decoders
		PARAFACDA+LDA and BTTDA+LDA evaluated on 2 motor imagery datasets.
		Scores for other decoders were taken from \textcite{Chevallier2024}.
    Accuracy is listed for BNCI2014-001, and \ac{rocauc} for BNCI2014-004.
    BTTDA outperforms HODA and PARAFACDA for BNCI2014-001 but does not reach
    performance comparable to the current state-of-the-art.
    For BNCI2014-004, PARAFACDA and BTTDA perform
    approximately on par with HODA and the current state-of-the art.}
	\label{tab:mi-score}
\end{table}
For the BNCI2014-001 dataset, HODA+LDA and our proposed decoders do not reach
satisfactory performance compared to the comparison methods, yet both
PARAFACDA+LDA and BTTDA+LDA improve upon HODA+LDA, and for BTTDA+LDA this
difference is significant ($p=0.0269$).
For the BNCI2014-004 dataset, the performance gap with comparison methods is
smaller, but no significant differences were found between HODA+LDA,
PARAFACDA+LDA, and BTTDA+LDA.

@@

Nevertheless, when considering the relative improvement over the original
HODA model, our results show that BTTDA+LDA performs consistently on par or
higher than HODA+LDA, with performances of respectively $86.43>83.25$ and
$85.08>82.57$ for \ac{erp} datasets and $55.13>53.51$ and $80.60<80.88$ for
\ac{mi} datasets.
The classification score for BTTDA+LDA was also slightly
higher than that of
restricted PARAFACDA+LDA for the ERP datasets ($86.43>86.19$ and $85.06>84.85$
respectively) and one of the MI datasets ($55.13>54.34$ and $80.60<80.88$
respectively), but these results were not significant, and further studies with
more datasets and subjects should show whether this holds.
\Cref{fig:blocks} shows that there is added value in finding extra blocks
over the first HODA block.
While no proof is given, we notice that NMSE monotonically decreases, indicating that
since the training score keeps increasing, this suggests that
eventually all the variation in the signal will be explained by the model
while still extracting features that are maximally discriminant.
Eventually, the number of blocks will reach a point of diminishing validation
score returns, when adding extra features to the decision classifier increases
its risk of overfitting instead of adding extra useful discriminatory
information.
Together with the improved classification scores presented, these results
point to the potential of our more flexible model block-term or
its special PARAFAC-structured case over a Tucker-structured model given proper block and
rank selection.
Since the optimal ranks for HODA+LDA were also determined through
cross-validation, BTTDA+LDA can improve over the first HODA block.
If validation shows that this is not possible, such as is the case for
BNCI2014-004, the BTTDA model is truncated to the first HODA block and little
performance is lost.

### 2024-11-27 chapters/patients.tex

\Ac{eeg} was recorded at 1000 Hz using the Neuroscan Neuvo portable amplifier (Compumedics Neuroscan,
Australia) connected to a second laptop for registration.
The \ac{eeg} headset used 18 active AgCl electrodes (EASYCAP GmbH, Germany) placed on a cap
according to the international 10-20 layout.
Using electrolyte gel, electrode impedances were reduced below 10k$\Omega$.
Additionally, the \ac{eog} was recorded.

The \ac{eeg} was band-pass filtered between 0.5 and 16 Hz.
Bad channels were rejected using the RANSAC algorithm~\cite{Fischler1981}
and visual inspection.
Next, the \ac{eeg} was re-referenced to the average of mastoid electrodes TP9
and TP10, and \ac{ica} was performed to reject artifactual components based on
correlation with the \ac{eog} or by visual inspection.
Epochs were cut from -0.1 to 0.9s relative to stimulus onset, and no baseline
correction was performed in order to meet the assumptions.

@@

In this study, user comfort in the different conditions was not objectively
measured.
Instead, it was assumed that participants operated most comfortably in the free
\ac{vsa} condition.
To properly contextualize performance results, they should be coupled with
metrics evaluating the user's requirements
with a measures of usability, comfort and perceived effort, like the NASA Task Load
Index~\cite{Hart2006} and other metrics proposed in the user-centered design
framework for \acp{bci}~\cite{Kuebler2014}.
Performance might, after all, be traded off for user comfort.
Eye motor disability could also have been assessed more
objectively~\cite{FriedOken2020}, using, e.g.,
the Revised Coma Recovery Scale~\cite{Giacino2004} or the NSUCO
oculomotor exam~\cite{Maples1992}.

>>>>

\Ac{eeg} was recorded at 1000 Hz using the Neuroscan Neuvo portable amplifier (Compumedics Neuroscan,
Australia) connected to a second laptop for registration.
The \ac{eeg} headset used 18 active AgCl electrodes (EASYCAP GmbH, Germany) placed on a cap
according to the international 10-20 layout.
Using electrolyte gel, electrode impedances were reduced below 10 k$\Omega$.
Additionally, the \ac{eog} was recorded.

The \ac{eeg} was band-pass filtered between 0.5 and 16 Hz.
Bad channels were rejected using the RANSAC algorithm~\cite{Fischler1981}
and visual inspection.
Next, the \ac{eeg} was re-referenced to the average of mastoid electrodes TP9
and TP10, and \ac{ica} was performed to reject artifactual components based on
correlation with the \ac{eog} or by visual inspection.
Epochs were cut from -0.1 to 0.9 s relative to stimulus onset, and no baseline
correction was performed in order to meet the assumptions.

@@

In this study, user comfort in the different conditions was not objectively
measured.
Instead, it was assumed that participants operated most comfortably in the free
\ac{vsa} condition.
To properly contextualize performance results, they should be coupled with
metrics evaluating the user's requirements
with a measure of usability, comfort and perceived effort, like the NASA Task Load
Index~\cite{Hart2006} and other metrics proposed in the user-centered design
framework for \acp{bci}~\cite{Kuebler2014}.
Performance might, after all, be traded off for user comfort.
Eye motor disability could also have been assessed more
objectively~\cite{FriedOken2020}, using, e.g.,
the Revised Coma Recovery Scale~\cite{Giacino2004} or the NSUCO
oculomotor exam~\cite{Maples1992}.

### 2024-11-27 chapters/stbf_struct.tex
Layout chapter gaze independence

  \noindent\emph{This chapter, except
  \crefrange{tab:stbf-struct/accuracies/stbf-struct}{tab:stbf-struct/accuracies/xdawn-rg},
  was published as~\textcite{VanDenKerchove2022}.}

@@

	\subsubsection{Empirical covariance estimation}
	\label{sec:stbf-struct/methods/emp-cov}
	Earlier spatiotemporal beamformer studies~\cite{Wittevrongel2016,
		Wittevrongel2016a, Wittevrongel2017, Wittevrongel2017a} use the empirical
	covariance and inverse covariance calculated as follows:
	\begin{equation}
    \hat{\mat{C}}_\text{emp} =
		\frac{1}{N-1}\sum^{N}_{n=1}\mat{x}_n\mat{x}_in\intercal
		\label{eq:stbf-struct/emp-cov}
	\end{equation}
	\begin{equation}
    \widehat{\mat{C}^{-1}}_\text{emp} = \hat{\mat{C}}_\text{emp}^+
		\label{eq:stbf-struct/emp-inv-cov}
	\end{equation}
	The Moore-Penrose pseudoinverse, $^+$, ensures a solution exists when
  $\hat{\mat{C}}_\text{emp}$ is singular.
	\cref{fig:kronlda-covs}a and \cref{fig:kronlda-covs}b respectively show examples of the
	empirical estimators of the covariance and the inverse covariance matrices.
	The empirical estimator suffers from performance and
	stability issues if the number of epochs $N$ used or estimation is not much larger than the number of features $CS$~\cite{Stein1956,Khatri1987}.

@@

	The \ac{eeg} recorded at 2048 Hz was re-referenced off-line to the average of the mastoids.
	The reference electrodes were dropped from the analysis.
	Data were subsequently filtered between 0.5 Hz and 16 Hz using forward-backward
	filtering with a fourth-order Butterworth IIR filter.
	The \ac{eeg} signal was corrected for ocular artifacts using Independent Component
	Analysis (ICA) by rejecting components that correlated with the bipolar EOG
  channels vEOG and hEOG using iterative Z-score thresholding.
  Components with a Z-scored pearson correlation coefficient
  exceeding the other 3 times the standard deviation of the Z-scored
  correlation coefficients of other component's Z-scored were iteratively
  rejected until none that exceed the threshold remained.
  Finally, epochs were cut from 0.1s to 0.6s after stimulus onset.
	No baseline correction was performed since this affects the temporal covariance
	of the data, violating the Toeplitz structure assumption~\cite{Bijma2003}.

@@

  \begin{figure}
    \sffamily
    \sansmath
    \begin{subfigure}{\linewidth}
      \input{figures/stbf_struct/fpi-0.pgf}
      \caption{Results for 1, 2, and 5 trials using only the first block in each
      training fold for training.}
    \end{subfigure}
    \smallskip

@@

  \begin{figure}
    %\includegraphics[width=.66\linewidth]{figures/stbf_struct/accuracy.eps}
    \sffamily
    \sansmath
    \input{figures/stbf_struct/accuracy.pgf}
    \caption[Clasifier accuracy in function of available training data.]{%
      Accuracy of the different classifiers for all 21 subjects relative to the
			number of blocks available for training. One block consists of 135
			epochs and corresponds to 27 seconds of stimulation. Accuracies
			are shown for the evaluation settings averaging over 1, 2, and
			3 trials of testing stimuli.
			\cref{fig:stbf-struct/accuracy-ap} contains results for all numbers of trials.
			While \textsc{stbf-emp} is unstable when little training data
			are available,
			regularization of the covariance matrix (\textsc{stbf-shrunk} and
			\textsc{stbf-struct}) drastically improves performance.}
		\label{fig:accuracy}
	\end{figure}

@@

	\begin{figure}
		%\includegraphics[width=.66\linewidth]{figures/stbf_struct/training_time.eps}
    \sffamily
    \sansmath
    \input{figures/stbf_struct/training_time.pgf}
    \caption[Classifier training time.]{Training time boxplots of different
    classifiers and individual results for all training folds for all subjects
    using 32 channels and a sampling rate of 32 Hz. Using structured covariance estimation greatly reduces
    training time.}
		\label{fig:stbf-struct/training-time}
	\end{figure}

@@

  \fullpagefig{%
    \sffamily
    \sansmath
    \input{figures/stbf_struct/accuracy_all.pgf}
  }{%
    \caption[Accuracy averaging over different numbers of trials.]{%
      Accuracy of the different classifiers for all 21 subjects relative to the
		  number of blocks available for training. One block consists of 135
		  epochs and corresponds to 27 seconds of stimulation. Accuracies
		  are shown for the evaluation settings averaging over different
	    numbers of trials, ranging from 1 to 15.
    }
    \label{fig:stbf-struct/accuracy-ap}
  }

>>>>

  \publicationnote{This chapter, except
  \crefrange{tab:stbf-struct/accuracies/stbf-struct}{tab:stbf-struct/accuracies/xdawn-rg},
  was published as~\textcite{VanDenKerchove2022}.}

@@

	\subsubsection{Empirical covariance estimation}
	\label{sec:stbf-struct/methods/emp-cov}
	Earlier spatiotemporal beamformer studies~\cite{Wittevrongel2016,
		Wittevrongel2016a, Wittevrongel2017, Wittevrongel2017a} use the empirical
	covariance and inverse covariance calculated as follows:
	\begin{equation}
    \hat{\mat{C}}_\text{emp} =
		\frac{1}{N-1}\sum^{N}_{n=1}\mat{x}_n\mat{x}_i^\intercal
		\label{eq:stbf-struct/emp-cov}
	\end{equation}
	\begin{equation}
    \widehat{\mat{C}^{-1}}_\text{emp} = \hat{\mat{C}}_\text{emp}^+
		\label{eq:stbf-struct/emp-inv-cov}
	\end{equation}
	The Moore-Penrose pseudoinverse, $^+$, ensures a solution exists when
  $\hat{\mat{C}}_\text{emp}$ is singular.
	\cref{fig:kronlda-covs}a and \cref{fig:kronlda-covs}b respectively show examples of the
	empirical estimators of the covariance and the inverse covariance matrices.
	The empirical estimator suffers from performance and
	stability issues if the number of epochs $N$ used or estimation is not much larger than the number of features $CS$~\cite{Stein1956,Khatri1987}.

@@

	The \ac{eeg} recorded at 2048 Hz was re-referenced off-line to the average of the mastoids.
	The reference electrodes were dropped from the analysis.
	Data were subsequently filtered between 0.5 Hz and 16 Hz using forward-backward
	filtering with a fourth-order Butterworth IIR filter.
	The \ac{eeg} signal was corrected for ocular artifacts using Independent Component
	Analysis (ICA) by rejecting components that correlated with the bipolar EOG
  channels vEOG and hEOG using iterative Z-score thresholding.
  Components with a Z-scored pearson correlation coefficient
  exceeding  3 times the standard deviation of the Z-scored
  correlation coefficients of other components, were iteratively
  rejected until none that exceed the threshold remained.
  Finally, epochs were cut from 0.1s to 0.6s after stimulus onset.
	No baseline correction was performed since this affects the temporal covariance
	of the data, violating the Toeplitz structure assumption~\cite{Bijma2003}.

@@

  \begin{figure}
    \sffamily
    \begin{subfigure}{\linewidth}
      \input{figures/stbf_struct/fpi-0.pgf}
      \caption{Results for 1, 2, and 5 trials using only the first block in each
      training fold for training.}
    \end{subfigure}
    \smallskip

@@

  \begin{figure}
    %\includegraphics[width=.66\linewidth]{figures/stbf_struct/accuracy.eps}
    \sffamily
    \input{figures/stbf_struct/accuracy.pgf}
    \caption[Clasifier accuracy in function of available training data.]{%
      Accuracy of the different classifiers for all 21 subjects relative to the
			number of blocks available for training. One block consists of 135
			epochs and corresponds to 27 seconds of stimulation. Accuracies
			are shown for the evaluation settings averaging over 1, 2, and
			3 trials of testing stimuli.
			\cref{fig:stbf-struct/accuracy-ap} contains results for all numbers of trials.
			While \textsc{stbf-emp} is unstable when little training data
			are available,
			regularization of the covariance matrix (\textsc{stbf-shrunk} and
			\textsc{stbf-struct}) drastically improves performance.}
		\label{fig:accuracy}
	\end{figure}

@@

	\begin{figure}
		%\includegraphics[width=.66\linewidth]{figures/stbf_struct/training_time.eps}
    \sffamily
    \input{figures/stbf_struct/training_time.pgf}
    \caption[Classifier training time.]{Training time boxplots of different
    classifiers and individual results for all training folds for all subjects
    using 32 channels and a sampling rate of 32 Hz. Using structured covariance estimation greatly reduces
    training time.}
		\label{fig:stbf-struct/training-time}
	\end{figure}

@@

  \fullpagefig{%
    \input{figures/stbf_struct/accuracy_all.pgf}
  }{%
    \caption[Accuracy averaging over different numbers of trials.]{%
      Accuracy of the different classifiers for all 21 subjects relative to the
		  number of blocks available for training. One block consists of 135
		  epochs and corresponds to 27 seconds of stimulation. Accuracies
		  are shown for the evaluation settings averaging over different
	    numbers of trials, ranging from 1 to 15.
    }
    \label{fig:stbf-struct/accuracy-ap}
  }
